How to publish in Biological Sciences: A guide for the uninitiated

### 2021-04-27 1.06-authors.Rmd
Add files via upload
# first upload of book files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
]{krantz}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={How to publish in Biological Sciences: A guide for the uninitiated},
  pdfauthor={John Measey},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=Blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.33,0.33,0.33}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.61,0.61,0.61}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.06,0.06,0.06}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.5,0.5,0.5}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.27,0.27,0.27}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.27,0.27,0.27}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.06,0.06,0.06}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.14,0.14,0.14}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.06,0.06,0.06}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.27,0.27,0.27}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.43,0.43,0.43}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.5,0.5,0.5}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.5,0.5,0.5}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.5,0.5,0.5}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage[bf,singlelinecheck=off]{caption}
\usepackage[scale=.8]{sourcecodepro}

\usepackage{framed,color}
\definecolor{shadecolor}{RGB}{248,248,248}

\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.8}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\floatpagefraction}{0.75}

\renewenvironment{quote}{\begin{VF}}{\end{VF}}
\let\oldhref\href
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}

\makeatletter
\newenvironment{kframe}{%
\medskip{}
\setlength{\fboxsep}{.8em}
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\renewenvironment{Shaded}{\begin{kframe}}{\end{kframe}}

\usepackage{makeidx}
\makeindex

\urlstyle{tt}

\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother

\frontmatter
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{How to publish in Biological Sciences: A guide for the uninitiated}
\author{John Measey}
\date{2021-04-27}

\begin{document}
\maketitle

% you may need to leave a few empty pages before the dedication page

%\cleardoublepage\newpage\thispagestyle{empty}\null
%\cleardoublepage\newpage\thispagestyle{empty}\null
%\cleardoublepage\newpage
\thispagestyle{empty}

\begin{center}
To my son,

without whom I should have finished this book two years earlier
%\includegraphics{images/dedication.pdf}
\end{center}

\setlength{\abovedisplayskip}{-5pt}
\setlength{\abovedisplayshortskip}{-5pt}

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\listoftables
\listoffigures
\hypertarget{welcome}{%
\chapter*{Welcome}\label{welcome}}

Welcome to my guide on how to publish in the biological sciences. This guide is not here to tell you what to write, but to demystify what goes on in the world of scientific publishing, and more broadly in the academic context. My intention is to demystify publication in the biological sciences, so that readers become aware of what is happening once they have submitted a manuscript, and how to better interpret the decisions made by colleagues who are reviewing and editing your work.

Publishing has become vital for all academics, such that it is recognised that we inhabit a `publish or perish' academic landscape. Yet for some, the process appears to require little or no stress, while others are left in the cold. This book is meant to be a guide to those uninitiated members of the academic community, postgraduate students and early career researchers, to bring them up to speed with all the background information on publishing.

\hypertarget{why-read-this-book}{%
\section*{Why read this book}\label{why-read-this-book}}

Publishing a paper in an academic journal should simply consist of submitting a publication worthy manuscript. But having a working knowledge of publishing will enable you to make better decisions about what, where and how to submit manuscripts. In this book, I explain the many choices that exist for those wishing to submit a manuscript for publication in the Biological Sciences. I describe publication bias, and how this is evidenced by reviewers and editors. Impact Factors and how the desire to track the performance of academics has led to unethical practices and exploitation of science and scientists. This guide provides an ``everything you wanted to know about publishing but were afraid to ask'' approach for anyone who still feels that there is more they need to know about publishing that might get them onto an even footing.

\hypertarget{structure-of-the-book}{%
\section*{Structure of the book}\label{structure-of-the-book}}

This book is written in four parts.

Part 1 - Getting your manuscript ready for submission

Part 2 - Submission, reviews and reviewing, revising and resubmitting

Part 3 - Once your paper is published

Part 4 - Further challenges in academia

\hypertarget{why-a-guide-for-the-uninitiated}{%
\section*{Why `A guide for the uninitiated'?}\label{why-a-guide-for-the-uninitiated}}

I think that most people with doctorates would agree that a PhD is not awarded to people because they are particularly bright or smart. If you had to be a genius then I wouldn't have a PhD. Indeed, I don't consider myself to be particularly clever, but I worked very hard to get my PhD. I was hampered by the fact that I didn't know anything about the goals and aims of the academic process of working towards a PhD, so it took a lot more work, wasted time, and (let's not mince our words) real pain. The end product was a fraction of the potential that I could have achieved, if I had understood more about the process. If I had only had a guide to tell me what it was all about, I could have saved myself so much time and energy. In short, I feel that I was uninitiated, and this is the guide I wish that I had had.

So, this guide is my practical attempt to help you to get you up to speed in the world of academic publishing, specifically for the biological sciences. Too often, however, it's a nightmare journey of cul-de-sacs, and groping in the dark.

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}

There are a great many people that I need to thank. First and foremost are my students, past and present, who have inspired me to put together first the blog posts and then the book. It is because you wanted more that I put this together. I have also been a student, and have been inspired by colleagues around the world who have been exemplary advisors. This book contains lots of links to blogs and articles written and posted freely on the internet by others who also aim to demystify and help. I thank this greater academic community (especially \#academicTwitter) for sharing and inspiring. Thanks go to the many reviewers and editors who have taken their time to improve my writing. I am still learning. Lots of the text in this book has been improved by feedback from my students and postdocs. A special mention must go to my brother, Richard, who has hosted my lab website for more than a decade, and especially for saving blog posts from hacking attacks. Thanks also to my wife, Thalassa, who proofread many of the blog posts after I had published them late at night, so that I could correct them over breakfast in the morning. James Baxter-Gilbert, Jack Dougherty, Anthony Herrel, Allan Ellis, Lisa and Mark O'Connell, James Vonesh, Carla Wagener all read or commented on different aspects of the book. Thanks are also due to my colleagues at the Centre for Invasion Biology, the Department of Botany and Zoology, and Stellenbosch University. A special thanks to the librarians who have supported many of my more extreme rantings about publishers.

\begin{flushright}
John Measey\\
Cape Town
\end{flushright}

\hypertarget{author}{%
\chapter*{About the Author}\label{author}}

John Measey is Associate Professor of Biological Sciences at Stellenbosch University. He has authored or co-authored more than 200 peer reviewed scientific papers and book chapters, and five books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals. He has graduated more than 20 postgraduate students, and his blog on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the beautiful Western Cape, South Africa.

\hypertarget{contribute}{%
\section*{Do you have something to contribute?}\label{contribute}}

This book is written in bookdown \citep{xie2016bookdown} specifically to make it a `live project' that will be open to anyone who wants to contribute, improve, or use as the basis for your own book. The easiest way for readers to contribute content directly is through a \href{https://help.github.com/articles/about-pull-requests/}{GitHub pull requests}. At the repository for this book, you will find Rmd files for each chapter, and as a GitHub user, you can simply edit the Rmd file and submit the changes. If I am happy with the changes proposed, I will merge your content with that of the book and add your name to the \protect\hyperlink{acknowledge}{Acknowledgements}.

One of the amazing potentials for \href{https://bookdown.org/}{bookdown books} is that all the files for this book are hosted in a repository on \href{https://github.com/johnmeasey/How-to-publish-in-Biological-Sciences/tree/main}{Github}. You have the opportunity to fork this repository and write your own version for a different discipline, a different language or for a different region of the world. It is also my hope that this guide can grow to become a community of practice for those conducting PhDs in Biological Sciences. It will not be possible to cover every aspect of writing a PhD in Biological Sciences, but it may be that I have missed out ones that are very important to you. Equally, parts of what is currently written will become obsolete as new initiatives begin, and old problems are resolved. For this reason, this guide needs to be a `living document', and anyone who wants to provide feedback or contribute new sections is more than welcome. Please feel free to open an issue, or make a Pull Request if you spot a typo.

\hypertarget{creative-commons-license}{%
\section*{Creative Commons License}\label{creative-commons-license}}

How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License. You may freely share this content for non-commercial purposes, with a source credit to www.HowToWriteAPhD.org

\hypertarget{disclaimer}{%
\section*{Disclaimer}\label{disclaimer}}

Although I have tried to make the information is this book as accurate as possible, it is provided without any warranty. The author and publisher have neither liability nor responsibility to any person or entity related to any loss or damages arising from the information contained in this book.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(}
  \FunctionTok{head}\NormalTok{(iris), }\AttributeTok{caption =} \StringTok{\textquotesingle{}We must have the iris data and some R code, otherwise it would not seem right\textquotesingle{}}\NormalTok{,}
  \AttributeTok{booktabs =} \ConstantTok{TRUE}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tab:iris}We must have the iris data and some R code, otherwise it would not seem right}
\centering
\begin{tabular}[t]{rrrrl}
\toprule
Sepal.Length & Sepal.Width & Petal.Length & Petal.Width & Species\\
\midrule
5.1 & 3.5 & 1.4 & 0.2 & setosa\\
4.9 & 3.0 & 1.4 & 0.2 & setosa\\
4.7 & 3.2 & 1.3 & 0.2 & setosa\\
4.6 & 3.1 & 1.5 & 0.2 & setosa\\
5.0 & 3.6 & 1.4 & 0.2 & setosa\\
\addlinespace
5.4 & 3.9 & 1.7 & 0.4 & setosa\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{part1}{%
\part{Part 1: Getting ready for submission}\label{part1}}

The reasons for publishing your work\ldots{}

I assume that you will all want to publish the work from your thesis, but it's worth going through the points here about why publication is so important.

\begin{itemize}
\tightlist
\item
  If your work is unpublished, then it will not be used. Without use, all the work that you put into it falls away.
\item
  Your published work will become the foundation for future students and academics who are active in your field. This might well be in another country or on another continent.
\item
  Your work was most likely paid for by tax-payers in the country where you studied. Publishing this work is a way of handing back the value of what you found. If it is published Open Access, the public can also read it for themselves.
\item
  Other funders of your thesis work may have made publication a prerequisite of the funding criteria.
\item
  It is hoped that by publishing your work it will become more accessible to the scientific community.
\item
  By publishing your work you will find that both you, and your work, become known by an international community that may well invite you to participate in the academic process further \citep{marks2013misuse}.
\item
  Both inside and outside of science as a career, peer reviewed published papers are seen as an important accomplishment in addition to the production of your thesis.
\item
  The work that you have already put into your thesis (especially if you have followed the advice in this book!), will mean that you are close to having chapters that can be submitted as publications.
\item
  Your advisor and lab colleagues who helped with your thesis work may depend on your publications.
\item
  There is a lot of satisfaction to be had by seeing your work published. It has been argued that this depends on who you are and where you come from \citep{husemann2017publicationism}.
\item
  The satisfaction of having someone say how enjoyable or inspiring your work is to read will certainly not be as frequent as if you managed to publish a novel or newspaper column. But when it does happen, it will bring a smile to your face.
\end{itemize}

This is by no means an exhaustive list, but I hope it will give you some insight into the importance of publishing your results. To counter this assertion, please be aware that there is also research that suggests that publication driven science is not healthy, and that we need a new way to motivate science \citep[e.g.][]{stergiou2014impact}. Indeed, it has been suggested that the `publish or perish' mentality of academia has resulted in the retention (and even deterioration) of poor experimental design, and hence higher false positives in the behavioural sciences \citep{smaldino2016natural}. We will come back to this later in part 4. However, before you can join the debate about the future, I'd argue that you need to provide your credentials for the present.

Given that you are reading this book, I will assume that you plan to publish your work. However, there are still some basics that I think it is worth considering in this first part of the book:

\begin{itemize}
\tightlist
\item
  What is a journal for?
\item
  What is it possible to publish?
\item
  What is the journal Impact Factor?
\item
  Why are citations important?
\end{itemize}

\hypertarget{whatjournalfor}{%
\chapter{What is a scientific journal for?}\label{whatjournalfor}}

I think it's worth reflecting on why we have scientific journals, and what they are for. Primarily journals are a means of communication for the academic community.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  They pass on the findings of individuals and teams of academics from all over the world.
\item
  By having a date when they are published, together with the names of the authors, they record primacy (see \href{taxonomy.html}{taxonomic names}); that is who came up with the finding or idea first.
\item
  They attempt to register legitimacy by collating and integrating comments and concerns through \href{peerreview.html}{peer review}.
\item
  Lastly, they archive these findings so that in future people can build on the work.
\end{enumerate}

.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  There are so many scientists in the world publishing so many papers that it's not possible for all scientists to read everything. Today annual growth in scientific papers of 5.1\%, equivalent to a doubling time of 13.8 years \citep{bornmann2020growth}. Contrast this with the early days when there were only two journals and they published all of the studies that were being undertaken at the time. Over time there has developed a natural hierarchical system of what scientists will read. This is reflected in citations, and the simplest measure of journals is something called the \href{impactfactor.html}{Impact Factor} which is discussed in a chapter below.
\item
  It has been said that `authorship' is a relatively modern concept, emerging from the empiricism of England's middle-ages \citep[see][]{cronin2001hyperauthorship}. In our recent history, it is considered to be important for individuals to record who thought of what and when. From these authors, we give societal `author-ity.' This gives credit where it's due. In the big scheme of things of course it's not important who did it. We know from historical examples like Darwin's theory of natural selection and Wallace's very similar thoughts will merely be a product of many people who were thinking about these ideas at the time. Although certain authors may be `ahead of their time', the majority of thoughts and ideas that come around today are a product of their time. However, for individuals and their institutions it can be important to claim credit as this may translate into some monetary value (e.g.~with patents) or a prestige value. The regulated system of a taxonomy puts a lot of importance on the priority of who described what and when (you can read more about that in the chapter on \href{taxonomy.html}{taxonomic nomenclature}.
\item
  The system of editors and peer reviewers determining whether or not a manuscript possesses sufficient merit to be published is still regarded as the gold standard in science \citep{mayden2012peer}. As you will discover, it is often a very high bar to achieve. Of course both editors and peer reviewers are human and so the system is not perfect. We'll talk more about peer review in a \href{peerreview.html}{chapter below}.
\item
  Archiving the findings of scientists is perhaps one of the most important roles of publishers that we should be most concerned about. In my career I have seen the changes from strictly paper dissemination of scientific findings as it was for the past 350 years, to primarily electronic findings many of which and never printed by the majority of readers. We should be concerned about how long these records will last. If you have never thought about the longevity of data storage, then this is something that you should give some thought to. We all need to change our perspectives on long-term thinking as this impacts almost every societal function (see the \href{https://longnow.org/}{Long Now Foundation}).
\end{enumerate}

\hypertarget{impactfactor}{%
\chapter{What is the Impact Factor of a journal, and why is it so important?}\label{impactfactor}}

The impact factor of a journal relates to the number of times each publication from the journal gets cited in the two years preceding the date of the Impact Factor (IF) (Equation \eqref{eq:Impact-Factor}). Thus, if you are thinking of publishing in a journal that has an IF of 1 you might expect that In the two years following the publication of your article you may get one citation. But this is not necessarily true for your paper. As discussed elsewhere you might be very good at publicising your work and have it extensively cited. One or two extensively cited papers might even change the impact factor of the journal without too many publications per year. As IFs are calculated on average for all papers published in that journal. If on the other hand you are thinking of publishing your article in a journal that has an IF of 5, you might expect that your article will be cited 5 times more than if you published in the first journal.

It's a relatively simple calculation as seen in equation \eqref{eq:Impact-Factor}:

\begin{equation*} 
\frac{Sigma  citations  in  journal X for  year Y}{(Number of publications in journal X for year Y-1) + (Number of publications in journal X for year Y-2)}\left( \int_{a}^{x} f(u)\,du\right)=f(x)
\label{eq:Impact-Factor}
\end{equation*}

Because all citations for year Y are needed before the IF can be calculated for each journal, IF for the preceding two years is typically not released until June of Y+1.

Impact Factor is calculated by Thomson Reuters based on their Science Citation Index now called the \href{https://www.webofknowledge.com/}{Web of Science}. This means that if your journal is not even listed in the Web of Science then it will not have any Impact Factor. The \href{https://www.webofknowledge.com/}{Web of Science} is continually policing the quality of the journals but it includes and this means from time to time channels are excluded. This tends to happen at the lower end of the Impact Factor scale. But recently it happened to some very well known journals and there was a big stink. You can read more about this \href{https://retractionwatch.com/2020/07/13/stunned-very-confused-two-more-journals-push-back-against-impact-factor-suppression/}{here}.

Note that there is a large conflict of interest here. A publisher, Thomson Reuters, owns the \href{https://www.webofknowledge.com/}{Web of Science} and that's the means to decide whether or not a publication can get an Impact Factor. Is it more likely that journals published by Thomson Reuters will be included in the Web of Science? Previously the Science Citation Index was a not-for-profit organisation. Now the organisation that is used by many of our employers in their means of evaluating our effectiveness is owned by a private company. This is certainly cause for concern. There is a group of people who are trying to replace Impact Factor with a group of other metrics, so perhaps by the time you read this Impact Factor will no longer be relevant. If you're interested read more \href{https://www.nature.com/articles/d41586-019-01643-3}{here}.

Do remember that there are a number of other citation indices including Scopus and Google Scholar. Google Scholar doesn't currently calculate the Impact Factor for all journals, only the top 100.

\hypertarget{from-a-simple-score-to-a-way-of-life}{%
\section{From a simple score to a way of life}\label{from-a-simple-score-to-a-way-of-life}}

When IF was originally devised, it wasn't supposed to govern the lives of academics. It was a calculation for librarians to help them decide which journals to keep and which to ditch under ever constrained budgets (caused by publishers' ever increasing prices). But along the way, this very simple index is now considered by many people to be a measure of quality, prestige and even academic success \citep{garfield1999journal}. Many people have highlighted how wrong these beliefs are (see here and here), but the growing trouble is that not only have many academics been misled, but so have administrators responsible for hiring and promotions.

In a paper by McKiernan et al \citeyearpar{mckiernan2019use}, they found that this metric features in the guidelines of many university panels responsible for the fate of academics jobs and therefore lives. Worryingly, many of these institutions don't actually talk about what IF measures. Instead they equate it with values and qualities that it certainly does not represent. Thus you may find that your career is influenced by a simple metric that almost all who use it don't actually understand what it means. The undue influence on lives of scientists that IF has led directly to the San Francisco Declaration on Research Assessment (known as DORA) in which many institutions and publishers signed up to. You should read this very simple declaration and find out whether your institution is a signatory.

\hypertarget{what-can-you-do-if-you-publish-a-journal-with-high-if}{%
\section{What can you do if you publish a journal with high IF?}\label{what-can-you-do-if-you-publish-a-journal-with-high-if}}

Very high ranking journals for Impact Factor are publications like Cell, Nature, Science and PNAS. This is because these publications are read by a very great number of people, and so are widely cited. Articles that get published in them receive a lot of attention from the press and media. This results in the prestige that a hiring institution might be looking for. If your academics publish in this journal, your institution may well receive lots of positive publicity. In some countries, notably China, there may be a cash incentive towards publishing in a journal with a high IF \citep{quan2017publish}.

One frightening trend that we are seeing in biological sciences is that the higher the Impact Factor the more the journal will charge you to publish in it. At the time of writing in November 2020 nature has just announced that they will charge \$9500 to publish in the highly-ranked journal (see \href{openaccess.html}{part 5}. This is more money than it cost to publish in any other journal at the moment.

\hypertarget{why-is-it-so-important}{%
\section{Why is it so important?}\label{why-is-it-so-important}}

Academics are measured by their productivity but also about the quality of their output. Because there are so many different academic disciplines the bean pushers who administer us need some way of ranking academics against each other. This is why they use the Impact Factor of the journals that the academic publishing in order to determine the quality of their output. Even though there are other metrics of the actual quality of an academic, most administrators continue to cling to IF and their beliefs of what it stands for (see blog post here).

Some countries reward their academics if they publish in high ranking journals. This can result in a salary bonus. It may also help with promotion, getting tenure or even just getting an interview for a job. If you're going to publish and you want a career in academia then you need to be aware of Impact Factors and what they mean to different stakeholders.

Many people will complain that their particular sub-discipline has a range of very low ranking journals with low impact factors. Others complain that journals with high impact factors tend to be edited by an old boys club that facilitates the members. In some cases like PNAS this is certainly true.

\hypertarget{editors-try-to-increase-if}{%
\section{Editors try to increase IF}\label{editors-try-to-increase-if}}

It's important to remember that editors care about Impact Factor \citep[see][]{ioannidis2019user}. This means that if they think your paper will not garner the same or more citations in two years as the current Impact Factor they are likely to reject it often without even sending it out to review. Editors also have been known to manipulate impact factors and there are a number of established ways of doing this \citep[see][]{metze2010bureaucrats, martin2016editors}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Ask authors to cite publications from your journal published within the last 2 years.
\item
  Ask reviewers to suggest publications from your journal published within the last two years to authors on which their review is conducted.
\item
  Encourage submission papers from laboratories with high output and citation rates
\item
  Reject papers that are likely to have no citations. This effectively reduces the size of the denominator in the above equation
\item
  Publishing issues in January means they have a maximum period of the year to get cited. This is now being inflated to having issues published online well ahead of the January date all the time gathering citations.
\item
  Encourage review articles
\item
  Editorials that cite every paper in the journal. This tactic is frequently used in special issues.
\end{enumerate}

Just like any metric, Impact Factor is liable for abuse. You need to be aware of how IF is used and abused by many people in the academic community.

\hypertarget{push-back-against-if}{%
\section{Push back against IF}\label{push-back-against-if}}

One very simple way that you can push back is to calculate IF scores for your papers and show how they relate to the IF of the journal that you publish in. In this way you are simply comparing your actual citations in the 2 years after your paper is published with the mean for the journal. There is a good chance that you generally get more citations than the mean for the journal, and with some statistics you can convincingly show that your citations are significantly higher than the IF. For this to be true, you might need to help your work get cited. \href{citations.html}{That's the subject of another chapter}.

\hypertarget{citations}{%
\chapter{Citations and how to get them}\label{citations}}

Citations matter as they reflect the number of people who found your research useful. Getting your research cited is not always straightforward. We have already mentioned here about publishing your research in journals with higher Impact Factors. This will likely help you get your research read. But there are lots of other tactics and here I have produced a list although I imagine it is not exhaustive.

\hypertarget{media-release}{%
\section{Media release}\label{media-release}}

Some journals will want you to write a media release that they may use to promote your paper. You can also proactively \href{pressrelease.html}{write a press release} and send it to your Media office at your institution. Writing a release for the media is very different from most of the writing that we've talked about so far and you can find a guide in a chapter below.

\hypertarget{social-media}{%
\section{Social media}\label{social-media}}

Social media is another way of getting your article to the attention of more people. Twitter has been adopted by many scientists and tweeting out an article that gets retweeted by the right people can get you tens or even hundreds of thousands of views. This is likely to be far in excess of anything your article will get by passively sitting on the journal website.

\hypertarget{popular-articles}{%
\section{Popular articles}\label{popular-articles}}

\href{populararticle.html}{Writing popular articles} is a good way to ensure that your paper gets some media attention. There are outlets like \href{https://theconversation.com/}{The Conversation} that specialise in publishing popular articles written by academics. But you might also find that academic societies that you belong to have newsletters that you can contribute a popular article to about your paper. As long as the circulation of these newsletters is big it should enhance the numbers of people that eventually read about your work.

\hypertarget{self-citations}{%
\section{Self-citations}\label{self-citations}}

>>>>

# When should you be an author? {#authors}

Who should be an author is being increasingly regulated because of widespread abuse, including ‘honorary authors’ or excluding ‘ghost authors’. Back in the day there were people that used to add fictitious characters to their author line (including their dogs!) as a joke. These days we have such a thing as ORCID that attempts to register all authors, turning them into numbers ([see part 3](orcid.html)). The need for all this regulation is because publications have turned into a kind of currency for academics. And when there is currency involved, abuse quickly results in human systems followed by the need for regulation. Hence, one of the results of going transparent is that we all become registered numbers, and our dogs won't get authorship anymore. Perhaps this is just the loss of an age of innocence (at least on the part of our dogs!).

A lot of the literature on the subject of ghost and honorary authorships has come from the medical profession, perhaps as this profession is prone to ghost authorship via the pharmaceutical industry [@matheson2016ghostwriting], and honorary authorship from heads of large research groups [see @rennie1994authorship].

## Ghost authorship 
Ghost authorship is when people are not included in the author line although they have not contributed to the study [@matheson2016ghostwriting]. Note that ghostwriting outside science is usually when someone who has the talent to write, writes the ideas of someone who doesn’t for a fee, and in return the latter takes the author line. This may well happen in science (especially when you consider the world of translation companies), but this is not really the meaning that we are discussing here.

There are many potential sources for ghost authors, including past students whose thesis work is taken by unscrupulous advisors, and published without their inclusion. More commonly, I believe, is that those who contribute substantially are not included as authors for political reasons (they have fallen out with those who are the authors), or they are simply forgotten because they have moved away from the institution. Ghost authorship is a land of the disenfranchised. This is becoming increasingly prevalent in the world of contributions of data, which is also freely accessible. Some authors will take and use the data, only referencing the doi for where they obtained it. Others will include the original people who created the data as authors because they value their continued insight and input. Where the situation becomes very messy is when some people are included and others are excluded. This is my experience where a paper simultaneously contains both honorary and ghost authors. A study by Wisler et al [-@wislar2011honorary] found ghost authorship in medical publications at 7.9%, although I’d argue that their methods (contacting corresponding authors) mean that the real levels are likely much higher.

While the world of inclusion (see below) has a warm and friendly glow about it (everyone appears to benefit), exclusion is characterised by lack of information contact and reasoning. If you are excluded from a publication even when you have contributed, you will not be getting an email from the authors detailing their decision. You’ll be lucky if they even send you a copy once it’s published. Meanwhile, those who are included will remain in the loop.

## Honorary authorship
Honorary authorship happens when people who have not contributed meaningfully to a study are included in the author line. If publications are the currency of science then you can see how being added to other people's publications increases your apparent productivity. While this might sound surprising to you, you should know that does happen and might be more common than you think. The Wisler et al [-@wislar2011honorary] study found that honorary authorship was as high as 17.6% in medical publications. It’s worth noting that this may vary between disciplines as there are various traditions in some disciplines whereby the head of a large team may always be included as an author of a paper that emerges from the team, whether or not they were involved. In biological sciences, teams tend to be quite small with a single or rarely multiple Principal Investigators (PIs). This means that your PI will likely be directly involved with your research and therefore also an author. Imagine a very large team with multiple PIs working under a head who insists that they are an author on every publication. This could add up to hundreds of publications in a year (e.g. [Yuri T. Struchkov is currently credited with >1600 publications on Scopus](https://www-scopus-com.ez.sun.ac.za/authid/detail.uri?authorId=35427689100)), and such prolific authorship has been questioned [e.g. @rennie1994authorship]. However, there appear to be very different levels of what could be considered credible and what incredible [e.g. 25 papers a year: @wager2015too], and I respond to this, and the general question of how prolific authors are becoming, below.
